---
title: "Gapminder Report"
author: "Learn to code: seminar 4"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    code_folding: hide
---

---------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, cache = F, message = F, warning = F,
  fig.height = 3, fig.width = 5, fig.align = 'center'
  )
```

We're going to be analyzing some data from the `gapminder` package.

```{r}
# install.packages(c('tidyverse','tidylog','gapminder', 'ggiraph','patchwork','gganimate','ggrepel','modelr'))
library(tidyverse)
library(tidylog, warn.conflicts = F)
library(gapminder)
library(ggiraph)
library(ggrepel)
library(patchwork)
library(gganimate)
library(ggrepel)
library(modelr)

ggplot2::theme_set(theme_minimal())

```


```{r}
# assign in global env (look at env pane)
gapminder <- 
  gapminder::gapminder |>
  mutate(country = str_remove_all(country, "'"))
```


- `country` is the unit of observation;  
- `continent` is a grouping variable for `country`  
- `time` indicates the ordering of repeated measures on `country`  
- `lifeExp`, `pop`, `gdpPercap` are the demographics of interest  
- 60 years of data for each country  



---------------------------------------------------------

## Create a linear model for each country

Let's create some linear models and talk about how appropriate they are for modelling the increase in life expectancy.

```{r fig.height=3, fig.width=3}
rwanda <- gapminder |> 
  filter(country %in% 'Rwanda')

# fit a linear model
model <- lm(lifeExp ~ year, data = rwanda)

rwanda |> 
  ggplot(aes(year, lifeExp)) +
  geom_path() +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = 'linear model: life expectancy ~ year')

rwanda |> 
  add_predictions(model) |> 
  add_residuals(model) |> 
  ggplot(aes(year, resid)) +
  geom_hline(yintercept = 0, alpha = 0.2) +
  geom_path() +
  geom_point() +
  labs(title = 'linear model residuals')
```

Now, we want to build a linear model for each country. But we don't want to copy paste those 20 lines for each country (since there are `r gapminder |> dplyr::distinct(country) |> nrow()` countries).

So we'll create a function `gapminder_model` that subsets the data for each country, fits an lm, then creates our plots

```{r}
# linear model analysis for a selected country
gapminder_model <- function(selected){
  # subset data
  country <- gapminder |> 
    filter(country %in% selected)
  
  # fit a linear model
  model <- lm(lifeExp ~ year, data = country)
  
  # plot linear model
  a <- country |> 
    ggplot(aes(year, lifeExp)) +
    geom_path() +
    geom_point() +
    geom_smooth(method = 'lm') +
    labs(subtitle = 'life expectancy ~ year')
  
  # plot residuals
  b <- country |> 
    modelr::add_predictions(model) |> 
    modelr::add_residuals(model) |> 
    ggplot(aes(year, resid)) +
    geom_hline(yintercept = 0, alpha = 0.2) +
    geom_path() +
    geom_point() +
    labs(subtitle = 'residuals')
    
  # using the patchwork package to combine plots
  a + b & plot_annotation(title = selected)
}

gapminder_model('Nigeria')
```


```{r}
gapminder_model('Canada')
```

---

That's still too laborious! Let's use a more programmatic approach by nesting the data and mapping our function over the list of dataframes.

```{r}
nested_data <- 
  gapminder |> 
  select(continent, country, year, lifeExp) |> 
  group_by(continent, country) |> 
  nest(data = c(year, lifeExp)) |> 
  ungroup()
```

```{r}
models <- nested_data |> 
  mutate(
    model = map(data, ~lm(lifeExp ~ year, data = .x)),
    coef = map(model, ~broom::tidy(.x)),
    quality = map(model, ~broom::glance(.x)),
    preds = map2(model, data, ~broom::augment(.x, .y))
  ) |> 
  glimpse()
```

## Model coeficients

```{r}

models |> 
  unnest(coef) |> 
  filter(term == 'year') |> 
  mutate(country = fct_reorder(country, estimate)) |> 
  ggplot(aes(
    x = continent,
    y = estimate, 
    ymin = estimate - std.error,
    ymax = estimate + std.error
  )) +
  geom_pointrange(alpha = 0.5, size = 0.25,
                  color = 'cornflowerblue',
                  position = position_jitter(width = 0.35)) +
  geom_boxplot(fill = NA, outlier.shape = NA) +
  labs(title = 'Increase in life expectancy per year (linear model)')
```

## Model quality

```{r fig.height=3, fig.width=4}
models |> 
  unnest(quality) |> 
  ggplot(aes(x = r.squared, color = continent, fill = continent)) +
  geom_histogram(color = 'white') +
  geom_rug(alpha = 0.5) 
```


---------------------------------------------------------

## Multiple plots with facet_*

```{r fig.height=2.9, fig.width=6}
#' life exp timeseries: trends for each continent
plt_life_exp <-
  gapminder |>
  ggplot(aes(year, lifeExp)) +
  geom_path(aes(group = country), alpha = 0.12) +
  geom_smooth(method = 'loess', se = F) +
  facet_grid(~continent) +
  scale_x_continuous(breaks = c(1960, 1980, 2000)) +
  theme(axis.text.x = element_text(size = 6))

plt_life_exp
```


---------------------------------------------------------

## How to best tell the story of this data?

Let's switch gears and talk about how to effectively visualize the trends in life expectancy and GDP over time.

```{r fig.height=5.5, fig.width=6}
## lifeExp vs gdp: all years
p1 <- gapminder |>
  ggplot(aes(gdpPercap, lifeExp, color  = continent, size = pop,
             label = country)) +
  geom_point(aes(alpha = year)) +
  geom_path(aes(group = country), size = 0.2, alpha = 0.3) +
  scale_x_log10() +
  scale_alpha(range = c(0.3, 0.8))
p1
```

We could add features to explore the data through zooming, hovering, and clicking on the image with `plotly`. This is really easy because we can use the function `ggplotly` that works for most ggplot objects.

```{r fig.height=5.5, fig.width=6}
plotly::ggplotly(p = p1)
```

If we were limited to static images, we might simplify the data, or show a sequence of slices.

```{r fig.height=6, fig.width=6}
# lifeExp vs gdp trajectories between 1952-2007
gapminder |>
  ggplot(
    aes(
      x = gdpPercap, 
      y = lifeExp,
      group = country, 
      color = year,
      )) +
  geom_path(alpha = 0.3, size = 1) +
  facet_wrap(~continent, scales = 'fixed') +
  scale_x_log10(breaks = 10**(3:5), labels = c('1', '10', '100')) +
  scale_color_viridis_c(option = 'E') +
  labs(x = 'GDP per capita (thousands of dollars)',
       y = 'Life Expectancy')
```


```{r fig.height=6, fig.width=6}
# lifeExp vs gdp trajectories over 1952-2007
gapminder |>
  ggplot(
    aes(
      x = gdpPercap, 
      y = lifeExp,
      color = continent,
      size = pop
      )) +
  geom_point(alpha = 0.5, shape = 1) +
  scale_x_log10(breaks = 10**(3:5), labels = c('1', '10', '100')) +
  facet_wrap(~year) +
  labs(x = 'GDP per capita (thousands of dollars)',
       y = 'Life Expectancy')
```


---------------------------------------------------------

## Animated plots 

```{r fig.height=6, fig.width=6}
# 10 biggest countries to label in animation
biggest_countries <-
  gapminder |> 
  group_by(country) |> 
  summarise(pop= max(pop)) |> 
  slice_max(pop, n = 20) |> 
  pull(country)

# make an plot will all time points...
p <- gapminder |>
  ggplot(aes(gdpPercap, lifeExp, color = continent, size = pop)) +
  geom_point(alpha = 0.5) +
  ggrepel::geom_text_repel(
    data = gapminder |> filter(country %in% biggest_countries),
    aes(label = country),
    size = 2.5,
    max.iter = 100,
    max.overlaps = 250) +
  scale_x_log10()

# before slicing by year
p


```

### Create an animation gif 

This gif isn't easy to get to display within Rmarkdown, so try calling `animated` in the console to see the gif in RStudio's viewer pane! The 'Eval' option is set to false on this chunk so that the markdown doesn't create the animation where rendering the html.

```{r eval=FALSE}
# save as a gif
# specify `year` as the variable that changes between frames
# add a title so we can see the year for each frame
library(gganimate)
library(gifski) # you might need to install this!
animated <- p +
  transition_states(year,
                    transition_length = 2,
                    state_length = 1) +
  ggtitle('Gapminder data: {closest_state}') +
  enter_fade() +
  exit_shrink()

anim_save('gapminder.gif', animation = animated)
```
 

---------------------------------------------------------

## Combine plots with patchwork

```{r fig.height=6, fig.width=6}
## interactive plot --------------------------------------------------------

# use a function to make similar plots with a different variable
demographic_timeseries <- function(data, demographic){
  plt <- data |>
    ggplot(aes(year, {{demographic}},  label = country)) +
    scale_x_continuous(breaks = c(1960, 1980, 2000)) +
    geom_line_interactive(
      aes(group = country, data_id = country,
          tooltip = country, hover_css = "fill:none;"),
      color = 'darkgray',
      alpha = 0.3
    ) +
    geom_smooth(data, mapping = aes(year, {{demographic}}, group = NULL),
    method = 'loess', se = F) +
    facet_grid(~continent) +
    labs(x = NULL) +
    theme(axis.text.x = element_text(size = 5))
  return(plt)
}

# call our function on each variable to make plots
plots <- list(
  pop = gapminder |> demographic_timeseries(pop) + scale_y_log10(),
  gpd = gapminder |> demographic_timeseries(gdpPercap) + scale_y_log10(),
  lifexp = gapminder |> demographic_timeseries(lifeExp)
)

# combine plots into three-part panel
demo_panel <- patchwork::wrap_plots(plots, ncol = 1)
demo_panel
```


---------------------------------------------------------
