---
title: "Learn to code:<h2>effective and accessible visualizations with ggplot2</h2><hr>️"
author: "Jason Moggridge"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: [default, 'custom2.scss']
    transition: fade
    background-transition: fade
    incremental: true
    echo: true
editor: source
---

## What is ggplot2?

<hr>

-   tidyverse graphics package

-   'grammar of graphics' syntax

-   sensible defaults

-   mostly customizable

-   extensible framework

```{r setup, echo=FALSE}
library(ggplot2)
theme_set(
  theme_bw() + 
  theme(
    text = element_text(size = 20), 
    strip.background = element_blank(),
    legend.position = 'top',
    plot.background = element_blank()
  ))
```

## The grammar of graphics

<hr>

-   data variables are mapped to aesthetics (`aes`)

-   plots are built in layers

## 'Aesthetic mapping' concept

<hr>

-   **Aesthetics** are visual channels to convey information.

-   How we link **variables** to aesthetics defines how data are displayed.

-   Different **layers** can have different aesthetic **mappings**

## Aesthetics

<hr>

-   x, y - position on coordinate plane
-   size
-   color, fill, alpha
    -   \*accessibility, limited bandwidth
-   shape or line-type
    -   \*discrete, limited bandwidth

## geom\_\*()

<hr>

These layers add shapes that display the data.

-   points -\> `geom_point`

-   lines -\> `geom_line`

-   bars -\> `geom_bar` or `geom_col`

-   boxplots, errorbars, pointranges, beeswarms, tiles, density, ...

## Example: 'iris'

<br>

Built-in data is great for learning / showing examples

```{r echo = T}
iris |> str()
```

## Basic ggplot2 usage - a histogram

<hr>

::: columns
::: {.column width="40%"}
<br>

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1|2|3|4-6|9"
ggplot(
  data = iris,
  aes(x = Petal.Length)
  ) +
  geom_histogram()
```
:::

::: {.column width="55%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
ggplot(
  data = iris,
  aes(x = Petal.Length)
  ) +
  geom_histogram()
```
:::
:::

## Basic ggplot2 usage - stacked histogram

<hr>

::: columns
::: {.column width="40%"}
<br>

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|4|6-9"
ggplot(
  data = iris,
  aes(x = Petal.Length,
      fill = Species)
  ) +
  geom_histogram(
    color = 'white',
    position = 'stack'
    )
```
:::

::: {.column width="55%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
ggplot(
  data = iris,
  aes(x = Petal.Length,
      fill = Species)
  ) +
  geom_histogram(
    color = 'white',
    position = 'stack'
    )
```
:::
:::

## Petal dimensions scatter

<hr>

::: columns
::: {.column width="35%"}
<br>

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1|2|3|4-6|9"
p <- ggplot(
  data = iris,
  aes(
    x = Petal.Length, 
    y = Petal.Width, 
    color = Species
    )
  ) +
  geom_point()
p
```
:::

::: {.column width="65%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
p <- ggplot(
  data = iris,
  aes(
    x = Petal.Length, 
    y = Petal.Width, 
    color = Species
    )
  ) +
  geom_point()
p
```
:::
:::

## Add more layers with '+'

<hr>

New layers inherit **aes** set in ggplot() but these can be added or over-ridden.

::: columns
::: {.column width="40%"}
<br>

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1|2-6"
p +
  geom_smooth(
    aes(fill = Species),
    alpha = 0.15,
    method = 'lm'
  )
```
:::

::: {.column width="55%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
p +
  geom_smooth(
    aes(fill = Species),
    alpha = 0.15,
    method = 'lm'
  )
```
:::
:::

## facet\_\*()

Split plots into multiple panels based on variable levels

::: columns
::: {.column width="35%"}
<br>

```{r}
#| echo: true
#| eval: false
p + facet_grid(~Species)
```
:::

::: {.column width="65%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
p + facet_grid(~Species)
```
:::
:::

## scale\_\*()

Alter the properties of the aesthetic space with scale\_\*() functions

::: columns
::: {.column width="45%"}
<br>

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|11|8,9"
p <- p + 
  scale_color_viridis_d(end = 0.9)
p
```
:::

::: {.column width="50%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
p <- p + scale_color_viridis_d(end = 0.9)
p
```
:::
:::

## labs()

Add titles and labels

::: columns
::: {.column width="45%"}
<br>

```{r}
#| echo: true
#| eval: false
p <- p + 
  labs(
    x = 'Length',
    y = 'Width',
    title = 'Iris petal dimensions'
  )
p
```
:::

::: {.column width="50%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
p <- p + 
  labs(
    x = 'Length',
    y = 'Width',
    title = 'Iris petal dimensions'
  )
p
```
:::
:::

## theme\*()

-   Select preset themes by adding `+ theme_*()`

-   Or override any parameter with `+ theme(...)`

-   Make theme\_\<your_name\>() by creating a function that returns a `theme()` list

## theme()

```{r}
p + theme(
  plot.title = element_text(family = 'Tourney', color = 'midnight blue'),
  axis.text = element_text(family = 'IBM Plex Sans', size = 11),
  plot.title.position = 'plot',
  legend.position = 'right'
)
```

## Extensions

::: columns
::: {.column width="45%"}
-   Easily combine multiple ggplot2s into complex layouts with `patchwork`

-   Make plots interactive using `ggiraph` on ggplot2
:::

::: {.column width="50%" style="padding-left: 0.5em;"}
```{r fig.height=6}
#| echo: false
plt <- iris |> 
  dplyr::mutate(tooltip = stringr::str_glue(
    "Iris {Species}
    Petal length: {Petal.Length}
    Petal width: {Petal.Width}
    "
  )) |> 
  ggplot(
    aes(
      Petal.Length, 
      Petal.Width, 
      color = Species,
      tooltip = tooltip
      )) +
  ggiraph::geom_point_interactive(
    alpha = 0.65,
    size = 2
    ) +
  scale_color_viridis_d(end = 0.9)

ggiraph::ggiraph(ggobj = plt, tooltip_extra_css = "font-family: Lobster; background-color: white;")

```
:::
:::

## ggplot2 art

```{r fig.height=6, fig.width=6, fig.align='center'}
#| echo: false
library(tidyverse)
p <- seq(-3, 3, by = .01) %>%
  expand_grid(x = ., y = .) |> 
  ggplot(aes(
    x = 1 - x - sin(y ^ 2),
    y = 1 - y - cos(x ^ 2),
    color = sin(x ^ 2)
  )) +
  geom_point(
    alpha = .45,
    shape = 15,
    size = 0,
    show.legend = F
  ) +
  scale_color_viridis_c(option = 'B') +
  coord_polar() +
  theme_void() 
p
```

## Code

```{r}
#| eval: false
library(tidyverse)
p <- seq(-3, 3, by = .01) %>%
  expand_grid(x = ., y = .) |> 
  ggplot(aes(
    x = 1 - x - sin(y ^ 2),
    y = 1 - y - cos(x ^ 2),
    color = sin(x ^ 2)
  )) +
  geom_point(
    alpha = .2,
    shape = 15,
    size = 0,
    show.legend = F
  ) +
  scale_color_viridis_c(option = 'B') +
  coord_polar() +
  theme_void() 
```

# Let's try it out
